### 2011-11-06 doc/pldi2012/highlevel.tex
More modification to the high level runtime section

\section{High-Level Runtime} 
\label{sec:highlevel}
In this section we describe a high level runtime system
that is designed to implement the Legion programming
and execution models while remaining machine agnostic.  
The high level runtime system operates on top of the 
low-level runtime which hides machine specific details
and is described in detail in section \ref{sec:lowlevel}.

The biggest challenge to implementing the high level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in solving this problem, the high level runtime employs two useful 
scheduling techniques.  The first technique was proved
in section \ref{sec:exec}: all scheduling decisions for a
task can be made locally in the context in which the
task was created.  The second technique is a {\em deferred execution model} 
that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies.  A
deffered execution model decouples the issuing of operations
from when they are actually performed.  This model is supported
by features in the low-level runtime that are discussed in section \ref{sec:lowlevel}.

@@

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  When
a task is registered with the high level runtime the task call includes
information about both the logical regions that the task will require
when it is executed as well as the read-write privileges and coherence properties
for each of the regions.  The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same context.  Dependence detection only needs to be performed at
the scope of an enclosing context as was proved in section \ref{sec:exec}.

@@

in section \ref{subsec:cohdep}.  We then describe the algorithm that
is used to perform dependence detection in section \ref{subsec:depdetect}.

@@

analysis is performed for the circuit simulation from section \ref{sec:ex}.

@@

When traversing a region node, we begin by checking if there are any tasks actively using
the node.  We determine this using the pair-wise dependence analysis described 
in section \ref{cohdep} to detect dependences between the region we are looking for and the
region represented by the current node.  For each dependence that is found, we record that the previous
task must have completed prior to the start of the task we are registering.

@@

In an out-of-order processor this problem is solved by register-renaming which maps logical
registers to a larger set of physical registers.  Our solution employs a similar technique
that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
is a place holder for the set of physical instances which a task will reference.  We call these
abstract instances because we may not know the actual physical instances until a task's predecessors
are mapped as described in section \ref{subsec:mapping}.  For each region in
a task we record the source and destination abstract instances that correspond to the physical
instances from which a task will read its input data and then place its output data.
%When the task is mapped as described in section \ref{subsec:mapping} we will update these
%abstract instances with the actual physical instances chosen.

@@

\subsubsection{Circuit Execution Example}
\label{subsec:cirdependence}
To illustrate concretely how dependence analysis works, we describe the steps for
performing dependence analysis for running circuit example.  When the {\tt simulate\_circuit}
task begins executing there are two region trees rooted at the logical
regions for {\tt all\_nodes} and {\tt all\_wires}.  For conciseness we will only
show the region tree for {\tt all\_nodes}.  At the start of the task, there exists
a single abstract instance at the root node representing the valid set of physical
instances from which sub tasks will be allowed to pull data.  The state of the
system at this point can be observed in figure~\ref{sfig:mapping_fig:cnc}.

When each of the {\tt compute\_new\_current} tasks are issued they must perform dependence
analysis for two node regions.  In the {\tt private\_node} case they will first open
up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub partition
for each of their subregions.  Since all of the private regions use a different subregion
of a disjoint partition for the private node accesses, no dependence is detected.  Similarly,
when all the dependence analysis is performed for the {\tt p\_ghost\_nodes} subregions,
each of the necessary partitions are opened.  There is no dependence with any of the private
nodes since all the ghost nodes traverse a different sub region of the disjoint partition
{\tt p\_nodes\_pvs}.  There is also no dependence between any tasks even though the
{\tt p\_ghost\_nodes} partition is aliased because all the ghost accesses are read-only.
The state of the region tree after dependence analysis has been performed for the
{\tt compute\_new\_current} nodes can be seen in figure~\ref{sfig:mapping_fig:dc}.

In the second phase of the computation, all of the {\tt distribute\_charge} tasks
will be issued.  For these tasks they will first analyze traverse the region tree
for their subregions containing their private nodes.  The path to these regions are
already open, but when the dependence analysis arrives at the specific private node subregion it will
discover a dependence between the task and the {\tt compute\_new\_currents} task that operated
on that region previously.  This dependence will be recorded and the 
{\tt distribute\_charge} task will be made the new active task as it dominated the
previous task.  For the ghost node regions, the dependence analysis will discover that the
there is a dependence between each of the {\tt distribute\_charge} tasks and all of the
{\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
performing writes and the partition is aliased.  However, the runtime will detect this as
anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
create new valid abstract instances rather than use the same abstract instances as
each of the {\tt compute\_new\_current} tasks.  There are no dependences detected
between each of the {\tt distribute\_charge} tasks on the ghost regions because
the {\tt Simultaneous} coherence property permits multiple tasks to be using the same
partition at the same time even if the subregions they are accessing are aliased.  The
state of the region tree after dependence analysis for {\tt distribute\_charge} can
be seen in figure~\ref{sfig:mapping_fig:volt}.

For the final phase of the computation, all of the {\tt update\_voltage} tasks
are issued requiring their private node region and their shared node region.  Again
the private node paths are already open, but a WAR dependence is detected which
requires a new abstract instance to be created to avoid a dependence.  The
shared nodes present a more interesting case.  The first shared node to be registered
will detect a dependence with every prior task since it must open the shared partition
instead of the ghost partition.  A new abstract instance is created at the 
{\tt all\_shared} node and copies are recorded for each valid instance in the ghost partition
back to the new {\tt all\_shared} abstract instance.  After this is complete, the
ghost partition can be opened.  Each corresponding {\tt update\_voltage} task will record
a dependence on the copy operation to close ghost partition as it traverses the {\tt all\_shared}
node ensuring that there is no requirement on the ordering of {\tt update\_voltage}
tasks.  The state of the region tree after this phase can be seen in figure \ref{fig:circdep4}.

\subsection{Mapping}
\label{subsec:mapping}
After tasks have undergone dependence analysis, the tasks are placed on either a ready
queue to be mapped or waiting queue.  A task is placed on the ready queue if all the tasks with
which it registered a dependence have mapped.  If any of the task's dependences have
not mapped then the task is placed on the waiting queue.  The task also registers
itself with the task it depends on so that when the dependent task is mapped it will notify
the task on the waiting queue. Tasks are mapped based on the partial ordering created by 
their data dependences.  By mapping tasks based on data dependences we guarantee the 
semantics of the programming model, while still allowing tasks to execute out of order.

When the scheduler is invoked, it will pull a series of tasks off the ready queue, map them,
and then issue them to be executed by the low level runtime.  The process of mapping a task
consists of several different queries made by the runtime to an application level object
called a {\em mapper}.  A mapper is an object which has control over how both tasks and
data can be placed in the machine.  This idea is a dynamic instantiation of the idea
initially presented in \cite{Fatahalian06} of creating mapping files for specifying how
tasks and data were placed in a memory hierarchy.  Rather than rely on a static mapping
we opted for a dynamic interface that could make decisions both based on application specific
input data as well as the machine specifics both at start-up and throughout the execution
of an application.  The interface for a mapper is described in section \ref{subsec:mapinter}.

\subsubsection{Mapper Interface}
\label{subsec:mapinter}
The interface for a mapper is designed to give the programmer control over how data and tasks
are placed in the machine.  Every mapper must support the following function calls:

\begin{itemize}
\item {\tt rank\_initial\_region\_locations} - When a new logical region is created, the runtime
will ask the mapper to give a ranking of physical memories of where to place the inital instance.

\item {\tt rank\_initial\_partition\_locations} - When a new partition is created, the mapper
must provide a ranking of memory locations for each of the new sub partitions.

\item {\tt select\_initial\_processor} - Given a task to be mapped, select a processor to
send this task to for mapping.

\item {\tt target\_task\_steal} - select a processor to attempt task stealing.  This is explained
in more detail in section \ref{subsec:steal}.

\item {\tt permit\_task\_steal} - also covered in section \ref{subsec:steal}

\item {\tt map\_task\_region} - Given a specific task and the logical region to be mapped, select
a source physical instance for the data, and provide a ranking of memories in which to create
or reuse an existing physical instance.

\item {\tt rank\_copy\_targets} - Given a task and a set of valid instances, generate a set of
rankings for new memories in which to place new phsyical instances.  This operation is used to
manage the copy operations performed when closing a subtree as they allow the programmer to
eagerly migrate regions through the memory hierarchy as copy ups are occuring.

\item {\tt select\_copy\_source} - Given a set of currently valid source instances and destination
physical instance, select the instance from which a copy should be made.
\end{itemize}

@@

In order to make it easy for programmers to write programs, we provide an initial implementation 
of the mapper interface that uses simple heuristics to place task and data.  This makes it easy
for the programmer to get an initial program running, while allowing them to write their own
mapper to tune for application or machine specific scenarios at a later time.

Lastly, the runtime supports multiple mappers in use in the same application.  For each task 
call the programmer can choose a mapper to be used for the runtime calls made inside that
task.  This allows for the creation of library code which provides their own mappers indpendent
from any application level mappers that may be created.

\subsubsection{Task Stealing for Load Balance}
\label{subsec:steal}
In order to achieve load balancing in an application the runtime supports an interface
for task stealing.  Our task stealing algorithm is loosely based on the stealing algorithm
used by Cilk\ref{CILK95} with guidance from the mapper about how stealing should be
performed.  Tasks are initially placed on the ready queue on the processor in which they
were registered.  However, a task can be moved to a different processor either by a mapper
explicitly choosing to relocate it, or by a steal request from another processor.

When a processor notices that it has no work on its ready queue, it will inquire into each
mapper asking for a target processor to attempt a steal via the {\tt target\_task\_steal}
call in the mapper interface.  A mapper can either respond with
a target processor, or indicate that no steal should be attempted.  The runtime will then
send steal requests to each processor on behalf of the specific mapper.

When a steal request is received at a processor it will determine the mapper(s) that generated
the list of steal requests.  For each mapper that issued a steal request, the runtime will
give to the local instance of that mapper a list of the tasks in the ready queue that that
specific mapper is responsible for mapping.  The mapper must then determine the set which
it is going to permit to be stolen if any.  Any tasks that the mapper marks as stolen are
sent back to the stealing processor.  By only allowing mappers to make steal requests about
the tasks they own, we are giving the mappers complete control over where tasks are mapped.
In addition, we are also giving the mappers control over how steal requests are made for
the case where the mapper has better knowledge about where additional work is than the 
runtime.  Finally, note that this mapping interface is sufficiently powerful to allow a user to
disable all stealing if desired, and to map tasks to explicit processors with no interference
from the runtime.

\subsection{Execution}
\label{subsec:execution}
Once task mapping is complete the high level runtime performs the necessary operations
to be issue the task to the low level runtime.  It first creates any new physical instances
for the regions which the task requires.  It then issues the necessary copies required
for the task to execute contignent on the tasks producing the source physical instances
have finished.  Then it issues the task itself to the low level processor contingent
on all the copies for the task haveing finished as well as the the set of tasks
on which the issuing task dependends having finished.  These dependences are all encoded
using event dependencies in the low level runtime.  Once all the necessary events
have triggered the task will be run.

Once a task starts running, the runtime is asked to create the region accessors for
the task.  The runtime investigates whether or not fast accessors can be created for
each of the regions that the task requires.  Fast accessors are described in more
detail in section \ref{sec:fastaccess}.  If fast accessors can be created for every
region in a task, then the runtime chooses to execute a task instantiated with fast
region accessor methods.  However, if any region can not be specialized into a fast
accessor the runtime defaults to a task instantiation that relies on {\tt get} and
{\tt put} operations to access regions.

The runtime will also return to the executing task a unique identifier describing the
context in which the task is executing.  All calls into the runtime require the task
to pass this context identifier which is what permits the runtime to know the context
for incoming calls.

\subsection{Task Completion}
\label{subsec:cleanup}
We define a task to have {\em completed} once the actual code for the task has completed.
We also define a task to have {\tt terminated} once all of the operations for a task
have finished.  We can only consider a task to be terminated once all of the regions that
the task accessed reflect all the changes made by a task and its child tasks.  However,
even after a task has completed it is possible for child tasks of that task
not to have been executed or even mapped.  We describe how we handle this scenario in
section \ref{subsec:restore}.  We also describe how physical instances are garbage
collected in seciont \ref{subsec:garbage}.

The runtime needs to know the mappings for
the child tasks as it must issue restoring copy operations so that the set of regions
that the task operated on are up to date with any data generated by child tasks.

\subsubsection{Restoring Copy Operations}
\label{subsec:restore}
An underlying assumption of the mapping process is that the resulting regions that
a task access will represent the data produced by that task.  However, this may not
be true if a task has child tasks which also modify subregions of the parent task.
In order to handle this issue the runtime must compute the necessary copy operations
from child task physical instances back to the parent task's physical instances.  To
compute these copy operations, the runtime first will wait until all the child tasks
have been mapped.  The runtime then issues a close operation to the each of the parent
task's region trees.  These close operations record all of the necessary copies to
restore the parent task's regions as well as the child tasks on which these copies
are contingent.  We call these copy operations at the end of a task {\em restoring copies}.
The runtime then issues these copies with the required dependences
and marks the task terminated once the restoring copy operations are complete.

\subsubsection{Garbage Collecting Physical Instances}
\label{subsec:garbage}
In addition to cleaning up task execution, the runtime must also deal with the problem
of garbage collecting physical instances of regions.  Physical instances must be
garbage collected since they are not visible to the application and would present
a correctness problem if deleting instances was an option for the mapper.  
Garbage collecting physical instances is a challenging problem
due to both the distributed environment of the machine and the deffered execution
model.  To solve this problem we employ a two-level reference counting scheme.  We
reference count both abstract instances and as well as physical instances.

An abstract instance maintains a count of the number of tasks that have referenced 
that abstract instance, but have not yet been mapped.  An abstract instance also tracks
whether or not it is open or closed.  An abstract instance is open if it is still
a valid abstract instance in a region tree.  An abstract instance is marked closed
once it is no longer a valid abstract instance and can therefore no longer be used
by any new tasks that are registered.

When a task is mapped it notifies an abstract instance that it has been used and can
decrement its reference count.  The task mapping also notifies the abstract instance
of the specific physical instance that was used and whether or not this use invalidates
previous physical instances of the abstract instance.  The abstract instance maintains
a reference count for each physical instance of it that has been made as well as
the set that are currently valid physical instances to be used for mapping.  When a 
task terminates, it notifies the abstract instances that it used to decrement the
reference counts for the task's specific physical instances.

There are two conditions under which a physical instance can be garbage collected.  The
first condition is when a physical instance is not a valid physical instance of
an abstract instance and its reference count is zero.  This condition occurs once no
more tasks are using a physical instance and it can longer be mapped by any task.  The
second condition is when an abstract instance is both closed and has a reference count
of zero as well as the physical instance's reference count being zero.  This handles the
case when no more tasks can be set to use an abstract instance, there are no more tasks
to map to an abstract instance, and there are no current tasks using the physical
instance.

>>>>

\section{High-Level Runtime} 
\label{sec:highlevel}
In this Section we describe a high level runtime system
that is designed to implement the Legion programming
and execution models while remaining machine agnostic.  
The high level runtime system operates on top of the 
low-level runtime which hides machine specific details
and is described in detail in Section \ref{sec:lowlevel}.

The biggest challenge to implementing the high level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in solving this problem, the high level runtime employs two useful 
scheduling techniques.  The first technique was proved
in Section \ref{sec:exec}: all scheduling decisions for a
task can be made locally in the context in which the
task was created.  The second technique is a {\em deferred execution model} 
that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies.  A
deffered execution model decouples the issuing of operations
from when they are actually performed.  This model is supported
by features in the low-level runtime that are discussed in Section \ref{sec:lowlevel}.

@@

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  When
a task is registered with the high level runtime the task call includes
information about both the logical regions that the task will require
when it is executed as well as the read-write privileges and coherence properties
for each of the regions.  The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same context.  Dependence detection only needs to be performed at
the scope of an enclosing context as was proved in Section \ref{sec:exec}.

@@

in Section \ref{subsec:cohdep}.  We then describe the algorithm that
is used to perform dependence detection in Section \ref{subsec:depdetect}.

@@

analysis is performed for the circuit simulation from Section \ref{sec:ex}.

@@

When traversing a region node, we begin by checking if there are any tasks actively using
the node.  We determine this using the pair-wise dependence analysis described 
in Section \ref{cohdep} to detect dependences between the region we are looking for and the
region represented by the current node.  For each dependence that is found, we record that the previous
task must have completed prior to the start of the task we are registering.

@@

In an out-of-order processor this problem is solved by register-renaming which maps logical
registers to a larger set of physical registers.  Our solution employs a similar technique
that maps logical regions to a larger set of {\em abstract instances}.  An abstract instance
is a place holder for the set of physical instances which a task will reference.  We call these
abstract instances because we may not know the actual physical instances until a task's predecessors
are mapped as described in Section \ref{subsec:mapping}.  For each region in
a task we record the source and destination abstract instances that correspond to the physical
instances from which a task will read its input data and then place its output data.
%When the task is mapped as described in section \ref{subsec:mapping} we will update these
%abstract instances with the actual physical instances chosen.

@@

\subsubsection{Circuit Execution Example}
\label{subsec:cirdependence}
To illustrate concretely how dependence analysis works, we describe the steps for
performing dependence analysis for circuit example from Section \ref{sec:ex}.  
For conciseness we will only show the node region tree.  At the start of the task, there exists
a single abstract instance at the root node representing the valid set of physical
instances from the parent task.

When each of the {\tt compute\_new\_current} tasks are issued they must perform dependence
analysis for two node regions.  In the {\tt private\_node} case they will first open
up the {\tt p\_nodes\_pvs} partition and then open up the {\tt p\_pvt\_nodes} sub partition
for each of their subregions.  We use a bold circle on an edge to indicate that it has been
opened.  Since all of the private regions use a different subregion
of a disjoint partition for the private node accesses, no dependence is detected.

Similarly, when the dependence analysis is performed for the {\tt p\_ghost\_nodes} subregions,
each of the necessary partitions are opened.  There is no dependence with any of the private
nodes since all the ghost nodes traverse a different subregion of the disjoint partition
{\tt p\_nodes\_pvs}.  There is also no dependence between any tasks even though the
{\tt p\_ghost\_nodes} partition is aliased because all the ghost accesses are read-only.
The state of the region tree after dependence analysis has been performed for the first
{\tt compute\_new\_current} task can be seen in figure~\ref{sfig:mapping_fig:cnc}.

In the second phase of the computation, all of the {\tt distribute\_charge} tasks
will be issued.  For these tasks they will first traverse the region tree
for the regions containing their private nodes.  The path to these regions are
already open, but when the dependence analysis arrives at the specific private node subregion it will
discover a dependence between the task and the {\tt compute\_new\_currents} task that operated
on that region previously.  This dependence will be recorded and the 
{\tt distribute\_charge} task will be made the new active task as it dominated the
previous task.

For the ghost node regions, the dependence analysis will discover that the
there is a dependence between each of the {\tt distribute\_charge} tasks and all of the
{\tt compute\_new\_current} tasks because the {\tt distribute\_charge} tasks are
performing writes and the partition is aliased.  However, the runtime will detect this as
anti-dependences which are resolved by forcing the {\tt distribute\_charge} tasks to
create new valid abstract instances rather than use the same abstract instances as
each of the {\tt compute\_new\_current} tasks.  There are no dependences detected
between each of the {\tt distribute\_charge} tasks on the ghost regions because
the {\tt Simultaneous} coherence property permits multiple tasks to be using the same
partition at the same time even if the subregions they are accessing are aliased.  The
state of the region tree after dependence analysis for first {\tt distribute\_charge} task can
be seen in figure~\ref{sfig:mapping_fig:dc}.

For the final phase of the computation, all of the {\tt update\_voltage} tasks
are issued requiring their private node region and their shared node region.  Again
the private node paths are already open, but a WAR dependence is detected which
requires a new abstract instance to be created to avoid a dependence.  The
shared nodes present a more interesting case.  The first shared node to be registered
will detect a dependence with every prior task since it must open the shared partition
instead of the ghost partition.  A new abstract instance is created at the 
{\tt all\_shared} node and copies are recorded for each valid instance in the ghost partition
back to the new {\tt all\_shared} abstract instance.  After this is complete, the
ghost partition can be opened.  Each corresponding {\tt update\_voltage} task will record
a dependence on the copy operation to close ghost partition as it traverses the {\tt all\_shared}
node ensuring that there is no requirement on the ordering of {\tt update\_voltage}
tasks.  The state of the region tree after the first update voltage task
has been registered can be seen in figure \ref{sfig:mapping_fig:volt}.

\subsection{Mapping}
\label{subsec:mapping}
After tasks have undergone dependence analysis, the tasks are placed on either a ready
queue to be mapped or waiting queue.  A task is placed on the ready queue if all the tasks with
which it registered a dependence have mapped.  If any of the task's dependences have
not mapped then the task is placed on the waiting queue.  The task also registers
itself with the task on which it depends so that when that task is mapped it will notify
the task on the waiting queue. Tasks are mapped based on the partial ordering created by 
their data dependences.  By mapping tasks based on data dependences we guarantee the 
semantics of the programming model, while still allowing tasks to execute out of order.

When the scheduler is invoked, it will pull a series of tasks off the ready queue, map them,
and then issue them to be executed by the low level runtime.  The process of mapping a task
consists of several different queries made by the runtime to an application level object
called a {\em mapper}.  A mapper is an object which has control over how both tasks and
data can be placed in the machine.  This idea is a dynamic instantiation of the idea
initially presented in \cite{Fatahalian06} of creating mapping files for specifying how
tasks and data are placed in a memory hierarchy.  Rather than rely on a static mapping
we opted for a dynamic interface that could make decisions both based on application specific
input data as well as the machine specifics both at start-up and throughout the execution
of an application.  %The interface for a mapper is described in Section \ref{subsec:mapinter}.

\subsubsection{Mapper Interface}
The mapper interface is a collection of eight function calls that allow the runtime to query
the application for information about how to map tasks are regions.  Mapper calls are made
from the runtime to ask where physical instances of regions should be place in the memory hierarchy,
which physical instances should be the source for copy operations, on which processors
tasks should be run, and how task stealing should be performed.
%\label{subsec:mapinter}
%The interface for a mapper is designed to give the programmer control over how data and tasks
%are placed in the machine.  Every mapper must support the following function calls:

%\begin{itemize}
%\item {\tt rank\_initial\_region\_locations} - When a new logical region is created, the runtime
%will ask the mapper to give a ranking of physical memories of where to place the inital instance.

%\item {\tt rank\_initial\_partition\_locations} - When a new partition is created, the mapper
%must provide a ranking of memory locations for each of the new sub partitions.

%\item {\tt select\_initial\_processor} - Given a task to be mapped, select a processor to
%send this task to for mapping.

%\item {\tt target\_task\_steal} - select a processor to attempt task stealing.  This is explained
%in more detail in Section \ref{subsec:steal}.

%\item {\tt permit\_task\_steal} - also covered in Section \ref{subsec:steal}

%\item {\tt map\_task\_region} - Given a specific task and the logical region to be mapped, select
%a source physical instance for the data, and provide a ranking of memories in which to create
%or reuse an existing physical instance.

%\item {\tt rank\_copy\_targets} - Given a task and a set of valid instances, generate a set of
%rankings for new memories in which to place new phsyical instances.  This operation is used to
%manage the copy operations performed when closing a subtree as they allow the programmer to
%eagerly migrate regions through the memory hierarchy as copy ups are occuring.

%\item {\tt select\_copy\_source} - Given a set of currently valid source instances and destination
%physical instance, select the instance from which a copy should be made.
%\end{itemize}

@@

In order to make it easy for programmers to write programs, we provide an initial implementation 
of the mapper interface that uses simple heuristics to place tasks and data.  For example, when
performing task stealing, the default mapper only encourages stealing tasks in groups that use
the same logical regions, which improves locality by moving tasks that use the same regions
together.  The default mapper implementation makes it easy for the programmer to get an 
initial program running.  As the programmer tunes the application, he can write mappers that
specialize for both application specific and/or machine specific parameters.

The high level runtime also supports the use of multiple mappers in the same application.  For each task 
call the programmer can choose a mapper to be used for the runtime calls made inside that
task.  This allows for the creation of library code which provides its own mappers indpendent
from any application level mappers.

\subsubsection{Task Stealing for Load Balance}
\label{subsec:steal}
In order to achieve load balancing in an application the runtime supports an interface
for task stealing.  Our task stealing algorithm is loosely based on the stealing algorithm
used by Cilk \cite{CILK95} with guidance from the mapper about how stealing should be
performed.  Tasks are initially placed on the ready queue on the processor in which they
were registered.  However, a task can be moved to a different processor either by a mapper
explicitly choosing to relocate it, or by a steal request from another processor's mappers.

When a processor notices that it has no work on its ready queue, it will inquire into each
mapper asking for a target processor to attempt a steal.  A mapper can either respond with
a target processor, or indicate that no steal should be attempted.  The runtime will then
send steal requests to each processor on behalf of each specific mapper.

When a steal request is received at a processor it will determine the mapper(s) that generated
the list of steal requests.  For each mapper that issued a steal request, the runtime will
give to the local instance of that mapper a list of the tasks in the ready queue that that
specific mapper is responsible for mapping.  The mapper must then determine the set which
it is going to permit to be stolen if any.  Any tasks that the mapper marks as stolen are
sent back to the stealing processor.  By only allowing mappers to make steal requests about
the tasks they own, we are giving the mappers complete control over where tasks are mapped.
In addition, we are also giving the mappers control over how steal requests are made for
the case where the mapper has better knowledge about where additional work is than the 
runtime.  Note that this mapping interface is sufficiently powerful to allow a user to
disable all stealing if desired, and to map tasks to explicit processors with no interference
from the runtime.

\subsection{Execution}
\label{subsec:execution}
Once task mapping is complete the high level runtime performs the necessary operations
to issue the task to the low level runtime.  It first creates any new physical instances
for the regions which the task requires.  It then issues the necessary copies required
for the task to execute contignent on the tasks producing the source physical instances
have finished.  Then it issues the task itself to the low level processor contingent
on all the copies for the task haveing finished as well as the the set of tasks
on which the issuing task dependends having finished.  These dependences are all encoded
using event dependencies in the low level runtime.  Events are described
in detail in Section \ref{subsec:events}.  Continuing the analogy of an out-of-order
processor, the execution phase is equivalent to issuing an instruction to an
execution unit, and having the execution unit defer the instruction's execution until all its
operands are ready.  The low-level runtime contains processors that act as execution
units that will run a task once all of the events required for the task to run
have occurred.

%Once a task starts running, the runtime is asked to create the region accessors for
%the task.  The runtime investigates whether or not fast accessors can be created for
%each of the regions that the task requires.  Fast accessors are described in more
%detail in Section \ref{sec:fastaccess}.  If fast accessors can be created for every
%region in a task, then the runtime chooses to execute a task instantiated with fast
%region accessor methods.  However, if any region can not be specialized into a fast
%accessor the runtime defaults to a task instantiation that relies on {\tt get} and
%{\tt put} operations to access regions.

The runtime also assigns a unique identifier describing the context
in which the task is executing.  All calls into the runtime require the task
to pass this context identifier which is what permits the runtime to know the context
for incoming calls for creating/destroying regions and launching sub-tasks.

\subsection{Task Completion}
\label{subsec:cleanup}
We define a task to have {\tt completed} once all of the task's regions reflect
the updated state of the task and all its sub-tasks.  This reflects an underlying assumption 
of the mapping process that the regions for which a task has priviliges will represent the data produced 
by that task.  However, this is not inherently true that a physical instance of a task reflects
this state after the task finishes executing.  If a task has child tasks these tasks may
still be executing when the parent task finishes the execution phase.

In order to handle this scenario the runtime must compute the necessary copy operations
from child task physical instances back to the parent task's physical instances.  To
compute these copy operations, the runtime must first wait until all the child tasks
have been mapped.  The runtime then issues a close operation to the each of the parent
task's region trees.  These close operations record all of the necessary copies to
restore the parent task's regions.  We call these copy operations at the end of 
a task {\em restoring copies}.
The runtime then issues these copies with the required dependences
and marks the task completed once the restoring copy operations are complete.
This is similar to the write-back phase of an out-of-order processor where store instructions cannot be
considered complete until the store effects the state of memory.

The last part of the completetion phase is garbage collecting physical instances.  Physical
instances are not a visible part of the programming model and therefore cannot be managed
directly by the programmer.  Having the mapper manage them would introduce correctness
issues.  We use a two-level reference counting scheme on both abstract instances and physical
instances to know when it is safe to reclaim physical instances.  Due to space constraints
we omit the details of this algorithm.

%Even after a task has completed it is possible for child tasks of that task not to have been 
%executed or even mapped.  This is similar
%to the write-back phase in an out-of-order processor where store instructions can only
%be considered terminated when they finish effecting the state of memory.

%The runtime needs to know the mappings for
%the child tasks as it must issue restoring copy operations so that the set of regions
%that the task operated on are up to date with any data generated by child tasks.

%\subsubsection{Restoring Copy Operations}
%\label{subsec:restore}
%\subsubsection{Garbage Collecting Physical Instances}
%\label{subsec:garbage}
%In addition to cleaning up task execution, the runtime must also deal with the problem
%of garbage collecting physical instances of regions.  Physical instances must be
%garbage collected since they are not visible to the application and would present
%a correctness problem if deleting instances was an option for the mapper.  
%Garbage collecting physical instances is a challenging problem
%due to both the distributed environment of the machine and the deffered execution
%model.  To solve this problem we employ a two-level reference counting scheme.  We
%reference count both abstract instances and as well as physical instances.

%An abstract instance maintains a count of the number of tasks that have referenced 
%that abstract instance, but have not yet been mapped.  An abstract instance also tracks
%whether or not it is open or closed.  An abstract instance is open if it is still
%a valid abstract instance in a region tree.  An abstract instance is marked closed
%once it is no longer a valid abstract instance and can therefore no longer be used
%by any new tasks that are registered.

%When a task is mapped it notifies an abstract instance that it has been used and can
%decrement its reference count.  The task mapping also notifies the abstract instance
%of the specific physical instance that was used and whether or not this use invalidates
%previous physical instances of the abstract instance.  The abstract instance maintains
%a reference count for each physical instance of it that has been made as well as
%the set that are currently valid physical instances to be used for mapping.  When a 
%task terminates, it notifies the abstract instances that it used to decrement the
%reference counts for the task's specific physical instances.

%There are two conditions under which a physical instance can be garbage collected.  The
%first condition is when a physical instance is not a valid physical instance of
%an abstract instance and its reference count is zero.  This condition occurs once no
%more tasks are using a physical instance and it can longer be mapped by any task.  The
%second condition is when an abstract instance is both closed and has a reference count
%of zero as well as the physical instance's reference count being zero.  This handles the
%case when no more tasks can be set to use an abstract instance, there are no more tasks
%to map to an abstract instance, and there are no current tasks using the physical
%instance.
